### 2017-11-26 paper/sections/1_introduction.tex
paper: spellcheck
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>
paper: addresses #45
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

Our analysis of the key-value activity generated by the ParSplice molecular
dynamics simulation demonstrates the need for more complex cache management
strategies. Baseline measurements show clear key access patterns and hot
spots that offer significant opportunity for optimization. We use the data
management and policy engine from the Mantle system to dynamically explore a
variety of techniques, ranging from basic algorithms and heuristics to
statistical models, calculus, and machine learning. While Mantle was originally
designed for distributed file systems, we show how the collection of
abstractions effectively decomposes the problem into manageable policies for a
different domain and service.  Our exploration of this space results in a
dynamically sized cache policy that, for our initial conditions, sacrifices
negligible performance while using only 28\% of the memory required by our
hand-tuned cache.

@@

Storage systems use software-based caches extensively to improve performance
but the policies that guide what data to evict and when to evict vary with the
use case. For example, caching file system metadata on clients and servers
reduces the number of remote procedure calls and improves the performance of
create-heavy workloads common in HPC~\cite{ren:sc2014-indexfs,
patil:fast2011-giga+}. But these the policies for when to evict and what data
to evict are specific to the application's behavior and the system's
configuration (hardware, settings, etc.) so a new use case may prove to be a
poor match for the selected caching
policy~\cite{xiao:socc15-shardfs,brandt:msst2003-lh,sevilla:sc15-mantle,
weil:sc2004-dyn-metadata, weil:osdi2006-ceph}. In this paper, we evaluate a
variety of caching policies using a data management language/policy engine and
arrive at a customized policy that works well for our target application.  This
process of trying general policies and quickly iterating to a customized
solution shows that our prototype can adapt to different workloads.

@@

Our target application, ParSplice~\cite{perez:jctc20150parsplice}, is
representative of these types of applications. ParSplice is a molecular
dynamics simulation that uses a hierarchy of caches and a single persistent
key-value store to store both observed minima across a molecule's equation of
motion (EOM) and the hundreds or thousands of partial trajectories calculated
each second during a parallel job.  The fine-grained data annotation
capabilities provided by key-value storage is a natural match for scientific
simulations like ParSplice. But these simulations rely on a mesh-based
decomposition of a physical region and result in millions or billions of mesh
cells, where each cell contains materials, pressures, temperatures and other
characteristics that are required to accurately simulate phenomena of interest.
Unfortunately, simulations of this size saturate the capacity and bandwidth
capabilities of a single node so we need more effective data management
techniques, such as cache management or load balancing across a cluster.

@@

The biggest challenge for ParSplice is properly sizing the caches in the
storage hierarchy.  The memory usage for a single cache that stores molecule
coordinates is shown in Figure~\ref{fig:cache-management}, where the thin solid
lines marked with \(\times\)'s are the existing configurations in ParSplice.
The default configuration uses an unlimited sized cache, shown by the ``No
Cache Management" line, but using this much memory for one cache is
unacceptable for HPC environments, where a common goal is to keep memory for
such data structures below 3\%\footnote{Anecdotally, we find this threshold
works well for HPC applications.  For reference, a 1GB cache for a distributed
file system is too large in LANL deployments.}. Since ParSplice deploys a cache
per 300 worker processes, large simulations will not scale.  Users can
configure ParSplice to evict data when the cache reaches a threshold but this
solution requires tuning and parameter sweeps; the ``Cache (too small)" curve
in Figure~\ref{fig:cache-management} shows how a poorly configured cache can
save memory but at the expense of performance, which is shown by the
annotation.  Even worse, this threshold changes with different initial
configurations and clusters so tuning needs to be done for all system
permutations.  Our dynamically sized cache, shown by the think line in
Figure~\ref{fig:cache-management}, detects key access patterns and re-sizes the
cache accordingly.  Without tuning or parameter sweeps, our solution saves more
memory than a hand-tuned cache with a negligible performance degradation
(within the baseline's standard deviation) and works for a variety of initial
conditions.

% What is Mantle
To design more flexible cache management policies, like our solution in
Figure~\ref{fig:cache-management}, we use the data management language and
policy engine from the Mantle paper~\cite{sevilla:sc15-mantle}.  While Mantle
was designed for the narrow purpose of filesystem metadata load balancing, this
paper provides evidence that the approach is more broad.  Specifically, the
collection of abstractions in Mantle provides a general control plane that
improves the performance of metadata access. So in this paper we refer to
Mantle as a policy engine that injects policies written in our data management
language directly into a running service, such as a file system or key-value
store.  Rather than co-designing a policy directly into the service, which
requires domain-specific knowledge to find and change policies, we expose the
policies in a general way so even developers unfamiliar with the domain can
deploy quickly deploy solutions.  We show that our framework:

@@

This last contribution is explored in Sections~\S\ref{sec:arch-specific}
and~\S\ref{sec:dom-specific}, where we try a range of policies from different
disciplines; but more importantly, in Section~\S\ref{sec:scope}, we conclude
that the collection of policies we designed for ParSplice's cache management
are very similar to the policies used to load balance metadata in the Ceph file
system (CephFS) suggesting that there is potential for automatically adapting
and generating policies dynamically.

>>>>

Our analysis of the key-value activity generated by the ParSplice molecular
dynamics simulation demonstrates the need for more complex cache management
strategies. Baseline measurements show clear key access patterns and hot spots
that offer significant opportunity for optimization. We use the data management
language and policy engine from the Mantle system to dynamically explore a
variety of techniques, ranging from basic algorithms and heuristics to
statistical models, calculus, and machine learning. While Mantle was originally
designed for distributed file systems, we show how the collection of
abstractions effectively decomposes the problem into manageable policies for a
different domain and service.  Our exploration of this space results in a
dynamically sized cache policy that, for our initial conditions, sacrifices
negligible performance while using only 28\% of the memory required by our
hand-tuned cache.

@@

Storage systems use software-based caches to improve performance but the
policies that guide what data to evict and when to evict vary with the use
case. For example, caching file system metadata on clients and servers reduces
the number of remote procedure calls and improves the performance of
create-heavy workloads common in HPC~\cite{ren:sc2014-indexfs,
patil:fast2011-giga+}. But the policies for when to evict and what data to
evict are specific to the application's behavior and the system's configuration
(hardware, settings, etc.) so a new use case may prove to be a poor match for
the selected caching
policy~\cite{xiao:socc15-shardfs,brandt:msst2003-lh,sevilla:sc15-mantle,
weil:sc2004-dyn-metadata, weil:osdi2006-ceph}. In this paper, we evaluate a
variety of caching policies using a data management language/policy engine and
arrive at a customized policy that works well for our target application.  This
process of trying general policies and quickly iterating to a customized
solution shows that our prototype can adapt to different workloads.

@@

Our target application, ParSplice~\cite{perez:jctc20150parsplice}, is
representative of applications that extensively use software-based caches.
ParSplice is a molecular dynamics simulation that uses a hierarchy of caches
and a single persistent key-value store to store both observed minima across a
molecule's equation of motion (EOM) and the hundreds or thousands of partial
trajectories calculated each second during a parallel job.  The fine-grained
data annotation capabilities provided by key-value storage is a natural match
for scientific simulations like ParSplice. But these simulations rely on a
mesh-based decomposition of a physical region and result in millions or
billions of mesh cells, where each cell contains materials, pressures,
temperatures and other characteristics that are required to accurately simulate
phenomena of interest.  Unfortunately, simulations of this size saturate the
capacity and bandwidth capabilities of a single node so we need more effective
data management techniques, such as cache management or load balancing across a
cluster.

@@

The biggest challenge for ParSplice is properly sizing the caches in the
storage hierarchy.  The memory usage for a single cache that stores molecule
coordinates is shown in Figure~\ref{fig:cache-management}, where the thin solid
lines marked with \(\times\)'s are the existing configurations in ParSplice.
The default configuration uses an unlimited sized cache, shown by the ``No
Cache Management" line, but using this much memory for one cache is
unacceptable for HPC environments, where a common goal is to keep memory for
such data structures below 3\%\footnote{Anecdotally, we find this threshold
works well for HPC applications.  For reference, a 1GB cache for a distributed
file system is too large in LANL deployments.}. ParSplice deploys a cache per
300 worker processes, so large simulations will not scale because they need
multiple caches.  Users can configure ParSplice to evict data when the cache
reaches a threshold but this solution requires tuning and parameter sweeps; the
``Cache (too small)" curve in Figure~\ref{fig:cache-management} shows how a
poorly configured cache can save memory but at the expense of performance,
which is shown by the annotation to the right.  Even worse, this threshold
changes with different initial configurations and cluster setups so tuning
needs to be done for all system permutations.  Our dynamically sized cache,
shown by the thick line in Figure~\ref{fig:cache-management}, detects key
access patterns and re-sizes the cache accordingly.  Without tuning or
parameter sweeps, our solution saves more memory than a hand-tuned cache with a
negligible performance degradation (within the baseline's standard deviation)
and works for a variety of initial conditions.

% What is Mantle
To design more flexible cache management policies, like our solution in
Figure~\ref{fig:cache-management}, we use the data management language and
policy engine from the Mantle paper~\cite{sevilla:sc15-mantle}.  While Mantle
was designed for the narrow purpose of file system metadata load balancing,
this paper provides evidence that the approach is more broad.  Specifically,
the collection of abstractions in Mantle provides a general control plane that
improves the performance of metadata access. So in this paper we refer to
Mantle as a policy engine that injects policies written in our data management
language directly into a running service, such as a file system or key-value
store.  Rather than co-designing a policy directly into the service, which
requires domain-specific knowledge to find and change policies, we expose the
policies in a general way so even developers unfamiliar with the domain can
quickly deploy solutions.  We show that our framework:

@@

This last contribution is explored in Sections~\S\ref{sec:arch-specific}
and~\S\ref{sec:dom-specific}, where we try a range of policies from different
disciplines; but more importantly, in Section~\S\ref{sec:scope}, we conclude
that the collection of policies we designed for cache management in ParSplice
are very similar to the policies used to load balance metadata in the Ceph file
system (CephFS) suggesting that there is potential for automatically adapting
and generating policies dynamically.

### 2017-11-26 paper/sections/2_parsplice-keyspace-analysis.tex
paper: addresses #45
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

\begin{figure}[t]
\noindent\includegraphics[width=0.5\textwidth]{figures/parsplice.png}\\
\caption{The ParSplice architecture has a storage hierarchy of caches and a
dedicated cache node (boxes) backed by a persistent database (DB). A splicer
(S) tells worker (W) to generate segments and workers employ tasks (T) for more
parallelization. We focus on the worker's cache (circled), which facilitates
communication and segment exchange between the worker and its tasks.
\label{fig:parsplice}}
\end{figure}

@@

Changing growth rates and temperature alters the size, shape, and locality of
the keyspace. Lower temperatures and smaller growth rates create
hotter keys with smaller keyspaces as many segments are generated in the same
set of states before the trajectory can escape to a new region of state space.
In Figure~\ref{fig:motivation-regimes}, the \(\Delta_1\) growth rate adds atoms
every 100K microseconds while the \(\Delta_2\) growth rate adds atoms every 1
million microseconds. So \(\Delta_2\) has a smaller growth rate resulting in
hotter keys (line on \(y1\) axis) and a smaller keyspace (dots on the \(y2\)
axis).  Values smaller than \(\Delta_2\)'s growth rate or a temperature of 400
degrees results in very little database activity because state transitions take
too long. Similarly, values larger than \(\Delta_1\)'s growth rate or a
temperature of 4000 degrees result in an equally meaningless simulation as
transitions are unrealistic.

@@

\subsubsection*{Scalability} Figure~\ref{fig:methodology-keyspace} shows the
keyspace size (text annotations) and request load (bars) after a one hour run
with a different number of workers (\(x\) axis). While the keyspace size and
capacity is relatively modest the memory usage scales linearly with the number
of workers. This is a problem if we want to scale to Trinitite's 6000 cores.
Furthermore, the size of the keyspace also increases linearly with the length
of the run.  Extrapolating these results puts an 8 hour run across all 100
Trinitite nodes at 20GB for the cache.  This memory utilization easily eclipses
the 3\% threshold we set earlier, even without factoring in the memory usage
from other workers.

\subsubsection*{An active but small keyspace}
Figure~\ref{fig:motivation-regimes} shows how ParSplice tasks read key-value
pairs from the worker's cache for two different initial conditions of
\(\Delta\), which is the rate that new atoms enter the simulation.  The line is
the read request rate (\(y1\) axis) and the dots along the bottom are the
number of unique keys accessed (\(y2\) axis).  This figure demonstrates that
small changes to \(\Delta\) can have a strong effect on the timing and
frequency with which new EOM minima are discovered and referenced. This finding
suggests that we need a flexible policy language and engine.

The bars in Figure~\ref{fig:methodology-keyspace} show \(50-100\times\) as many
reads (\texttt{get()}) as writes (\texttt{put()}).  Worker tasks read the same
key for extended periods because the trajectory can remain stuck in so-called
superbasins composed of tightly connected sets of states. In this case, many
trajectory segments with the same coordinates are needed before the trajectory
moves on.  Writes only occur for the final state of segments generated by
worker tasks; their magnitude is smaller than reads because the caches ignore
redundant write requests. The number of read and write requests are highest at
the beginning of the run when worker tasks generate segments for the same
state, which is computationally cheap (this motivates
Section~\S\ref{sec:dom-specific}).

\subsubsection*{Entropy increases over time} The reads per second in
Figure~\ref{fig:motivation-regimes} show that the number of requests decreases
and the number of active keys increases over time. The resulting key access
imbalance for the two growth rates in Figure~\ref{fig:motivation-regimes} are
shown in Figure~\ref{fig:methodology-keys}, where reads are plotted for each
unique state, or key, along the \(x\) axis. Keys are more popular than others
(up to \(5\times\)) because worker tasks start generating states with different
coordinates later in the run (this motivates Section~\S\ref{sec:dom-specific}).
The growth rate, temperature, and number of workers have a predictable effect
on the structure of the keyspace.  Figure~\ref{fig:methodology-keys} shows that
the number of reads changes with different initial conditions (\(\Delta\)), but
that the spatial locality of key accesses is similar ({\it e.g.}, some keys are
still \(5\times\) more popular than others).
Figure~\ref{fig:motivation-regimes} shows how entropy for different growth
rates has temporal locality, as the reads per second for \(\Delta_2\) looks
like the reads per second for \(\Delta_1\) stretched out along the time axis.
Trends also exist for temperature and number of workers but are omitted here
for space. This structure means that we can learn the key access patterns and adapt the
storage system accordingly (this motivates Section~\S\ref{sec:dom-specific}).

>>>>

\begin{figure}[t]
\noindent\includegraphics[width=0.5\textwidth]{figures/parsplice.png}\\
\caption{The ParSplice architecture has a storage hierarchy of caches (boxes) and a
dedicated cache process (large box) backed by a persistent database (DB). A splicer
(S) tells workers (W) to generate segments and workers employ tasks (T) for more
parallelization. We focus on the worker's cache (circled), which facilitates
communication and segment exchange between the worker and its tasks.
\label{fig:parsplice}}
\end{figure}

@@

@@

\subsubsection*{Scalability} Figure~\ref{fig:methodology-keyspace} shows the
keyspace size (text annotations) and request load (bars) after a one hour run
with a different number of tasks(\(x\) axis). While the keyspace size and
capacity is relatively modest the memory usage scales linearly with the number
of tasks. This is a problem if we want to scale to Trinitite's 3000 cores.
Furthermore, the size of the keyspace also increases linearly with the length
of the run.  Extrapolating these results puts an 8 hour run across all 100
Trinitite nodes at 20GB for the cache.  This memory utilization easily eclipses
the 3\% memory usage per node threshold we set earlier, even without factoring
in the usage from other workers.

\subsubsection*{An active but small keyspace}

Figure~\ref{fig:motivation-regimes} shows how ParSplice tasks read key-value
pairs from the worker's cache for two different initial conditions of
\(\Delta\), which is the rate that new atoms enter the simulation.  The line is
the read request rate (\(y1\) axis) and the dots along the bottom are the
number of unique keys accessed (\(y2\) axis).  The \(\Delta_1\) growth rate
adds atoms every 100K microseconds while the \(\Delta_2\) growth rate adds
atoms every 1 million microseconds. So \(\Delta_2\) has a smaller growth rate
resulting in hotter keys (line on \(y1\) axis) and a smaller keyspace (dots on
the \(y2\) axis).  Values smaller than \(\Delta_2\)'s growth rate or a
temperature of 400 degrees results in very little database activity because
state transitions take too long. Similarly, values larger than \(\Delta_1\)'s
growth rate or a temperature of 4000 degrees result in an equally meaningless
simulation as transitions are unrealistic.  This figure demonstrates that small
changes to \(\Delta\) can have a strong effect on the timing and frequency with
which new EOM minima are discovered and referenced. This finding suggests that
we need a flexible policy language and engine to explore these trade-offs.

The bars in Figure~\ref{fig:methodology-keyspace} show \(50-100\times\) as many
reads (\texttt{get()}) as writes (\texttt{put()}).  Worker tasks read the same
key for extended periods because the trajectory can remain stuck in so-called
superbasins composed of tightly connected sets of states. In this case, many
trajectory segments with the same coordinates are needed before the trajectory
moves on.  Writes only occur for the final state of segments generated by
tasks; their magnitude is smaller than reads because the caches ignore
redundant write requests. The number of read and write requests are highest at
the beginning of the run when tasks generate segments for the same state, which
is computationally cheap (this motivates Section~\S\ref{sec:arch-specific}).

\subsubsection*{Entropy increases over time} The reads per second in
Figure~\ref{fig:motivation-regimes} show that the number of requests decreases
and the number of active keys increases over time. The resulting key access
imbalance for the two growth rates in Figure~\ref{fig:motivation-regimes} are
shown in Figure~\ref{fig:methodology-keys}, where reads are plotted for each
unique state, or key, along the \(x\) axis. Keys are more popular than others
(up to \(5\times\)) because worker tasks start generating states with different
coordinates later in the run.  The growth rate, temperature, and number of
workers have a predictable effect on the structure of the keyspace.
Figure~\ref{fig:methodology-keys} shows that the number of reads changes with
different initial conditions (\(\Delta\)), but that the spatial locality of key
accesses is similar ({\it e.g.}, some keys are still \(5\times\) more popular
than others).  Figure~\ref{fig:motivation-regimes} shows how entropy for
different growth rates has temporal locality, as the reads per second for
\(\Delta_2\) look like the reads per second for \(\Delta_1\) stretched out
along the time axis.  Trends also exist for temperature and number of workers
but are omitted here for space. This structure means that we can learn the key
access patterns and adapt the storage system accordingly (this motivates
Section~\S\ref{sec:dom-specific}).

### 2017-11-26 paper/sections/3_methodology.tex
paper: spellcheck
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>
paper: addresses #45
Signed-off-by: Michael Sevilla <mikesevilla3@gmail.com>

%  \item background: CephFS implementation
To explore software-defined cache management, we use
the data management language and policy engine presented
in~\cite{sevilla:sc15-mantle}. The prototype in that paper was called Mantle
and it was originally built on the CephFS so administrators could control file
system metadata data management policies. We now refer to Mantle as a policy
engine that supports our data management language.  The basic premise is that
data management policies can be expressed with a simple API consisting of
``when", ``where", and ``how much". ``when" controls how aggressive or
conservative the decisions are; ``where" lets the policy control how
distributed or concentrated the data should be; and ``how much" controls the
amount of data that should be sent. There is also a ``load" policy that lets
administrators specify how to collapse many metrics into a single load metric
({\it e.g.}, \(2\times\texttt{cpu} + 3\times\texttt{memory usage}\)).

The succinctness of the API lets users inject multiple, possibly dynamic, policies. In
this work we focus on a single node, so the ``where" policy is not used. When
we move ParSplice to a distributed key-value store back-end, the ``where" policy
can be used determine which key-value pairs should be moved to which node.

@@

@@

% policies themselves
\subsubsection{Policies Written as Callbacks} administrators write policies in
Lua that are executed whenever the service needs to make a data management
decision. We chose Lua for simplicity, performance, and portability; it is a
scripting language with simple syntax, which allows administrators to focus on
the policies themselves; it was designed as an embeddable language, so it is
lightweight and does less type checking; and it interfaces nicely with C/C++.
The ``callback" box in Figure~\ref{fig:mantle} shows an example policy for the
``when()" callback in a distributed service, where the current server
(\texttt{whoami}) migrates load if it is has load (\texttt{>0.1}) and if its
neighbor server (\texttt{whoami + 1}) does not have load (\texttt{<0.1}). The
load is calculated using the metrics provided by the environment.

%of timestamp (ts), value pairs representing accesses over time types of
\subsubsection{Environment of Metrics} services can expose cluster metrics for
describing resource utilization and time series metrics for describing accesses
to some data structure over time. Table~\ref{table:metrics} shows how these
metrics are accessed from the policies written by administrators.

@@

\subsection{Integrating Mantle into ParSplice}

% env of metrics
Using Mantle cluster metrics, we expose cache size, CPU utilization, and memory
pressure of the worker node to the cache management policies. In
Section~\S\ref{sec:parsplice-keyspace-analysis} we only end up using the cache
size although the other metrics proved to be valuable debugging tools. Using
Mantle time series metrics, we expose accesses to the cache as a list of
\texttt{timestamp, keyid} pairs. In Section~\S\ref{sec:dom-specific}, we
explore a key access pattern detection algorithm that uses this metric.

>>>>

%  \item background: CephFS implementation
To explore software-defined cache management, we use the data management
language and policy engine presented in~\cite{sevilla:sc15-mantle}. The
prototype in that paper was called Mantle and it was originally built on CephFS
so administrators could control file system metadata data management policies.
We now refer to Mantle as a policy engine that supports our data management
language.  The basic premise is that data management policies can be expressed
with a simple API consisting of ``when", ``where", and ``how much". ``when"
controls how aggressive or conservative the decisions are; ``where" controls
how distributed or concentrated the data should be; and ``how much" controls
the amount of data that should be sent. There is also a ``load" policy that
lets administrators specify how to collapse many metrics into a single load
metric ({\it e.g.}, \(2\times\texttt{cpu} + 3\times\texttt{memory usage}\)).

The succinctness of the API lets users inject multiple, possibly dynamic, policies. In
this work we focus on a single node, so the ``where" policy is not used. When
we move ParSplice to a distributed key-value store back-end, the ``where" policy
will be used determine which key-value pairs should be moved to which node.

@@

Administrators write policies in
Lua that are executed whenever the service needs to make a data management
decision. We chose Lua for simplicity, performance, and portability; it is a
scripting language with simple syntax, which allows administrators to focus on
the policies themselves; it was designed as an embeddable language, so it is
lightweight and does less type checking; and it interfaces nicely with C/C++.

@@

\subsubsection{Environment of Metrics} services expose cluster metrics for
describing resource utilization and time series metrics for describing accesses
to some data structure over time. Table~\ref{table:metrics} shows how these
metrics are accessed from the policies written by administrators.

@@

% policies themselves
\subsubsection{Policies Written as Callbacks} the ``callback" box in
Figure~\ref{fig:mantle} shows an example policy for ``when()", where the
current server (\texttt{whoami}) migrates load if it is has load
(\texttt{>0.1}) and if its neighbor server (\texttt{whoami + 1}) does not have
load (\texttt{<0.1}). The load is calculated using the metrics provided by the
environment.

%of timestamp (ts), value pairs representing accesses over time types of
\subsection{Integrating Mantle into ParSplice}

% env of metrics
Using Mantle cluster metrics, we expose cache size, CPU utilization, and memory
pressure of the worker node to the cache management policies. In
Section~\S\ref{sec:parsplice-keyspace-analysis} we only end up using the cache
size although the other metrics proved to be valuable debugging tools. Using
Mantle time series metrics, we expose accesses to the cache as a list of
\texttt{timestamp, key} pairs. In Section~\S\ref{sec:dom-specific}, we
explore a key access pattern detection algorithm that uses this metric.
